---
title: 'Looking Deeper at Panel Data'
description: 'This chapter takes a deeper dive into the terms and choices you need to be aware of when working with panel data.'
---

## Panel Data Terminology

```yaml
type: VideoExercise
key: bd7bd16de8
lang: r
xp: 50
skills: 1
video_link: //player.vimeo.com/video/226207441
```


---

## Why use a lagged dependent variable?

```yaml
type: MultipleChoiceExercise
key: b9266032e1
lang: r
xp: 50
skills: 1
```

Think about what you know about causal inference. What might be an advantage for modeling your data with a lagged dependent variable; that is, a dependent variable measured at a time point after your explanatory variables were measured?

`@possible_answers`
- Letting time pass allows confounding factors to disappear.
- It gives you more time to formulate your hypothesis.
- It helps you more clearly establish the causal direction.
- A dependent variable depends on time, so dependent variables always happen after the independent variables.

`@hint`


`@pre_exercise_code`
```{r}

```

`@sct`
```{r}
msg1 = "Time isn't really a factor with determing what is a confounder. If an outside variable affects outcomes, it's a confounder, no matter when in the process it happens. Try again"
msg2 = "You should be testing a hypothesis you formed before you started, otherwise you aren't following the scientific method! Try again"
msg3 = "Well done"
msg4 = "You have labeled the dependent and independent variables based on your theory and knowledge of the research question, not based on the direction of the arrow of time. So try again. "
test_mc(correct = 3, feedback_msgs = c(msg1,msg2,msg3,msg4))
```

---

## Practice with lagged dependent variables

```yaml
type: NormalExercise
key: 6277c6d8ad
lang: r
xp: 100
skills: 1
```

Walter is the CEO of a large adult coloring books company. As part of a new advertising campaign, Walter wants to publish research which demonstrates that coloring in adult coloring books leads to reduced stress levels. 

When individuals purchase Walter's coloring books online, his company automatically sends a questionnaire asking about his sample's current stress levels. Although respondents reported that coloring in the adult coloring books helped reduce their feelings of stress, the sample's stress levels were about the same as a random sample of individuals who did not buy coloring books. 

Walter hypothesizes that the individuals who purchased the coloring books did not have different stress levels than the general population because they were administered the survey immediately following their purchase. The beneficial effects of using adult coloring books may take time to become noticeable. Therefore, Walter sent a follow-up survey regarding his original sample's stress levels a year later. Using the dataset, `Color`, and help Walter estimate the effect of adult coloring books on stress with his follow up survey.

`@instructions`
- 1) Reproduce Walter's initial analysis by estimating an OLS model that measures the immediate effect of adult coloring books (`books`) on stress (`stress1`), controlling for age (`age`) and gender (`sex`).
- 2) Estimate an OLS model that measures the effect of adult coloring books (`books`) on stress with the lagged dependent variable (`stress2`), controlling for age (`age`) and gender (`sex`).
- 3) Compare the two models to determine which measurement of stress (`stress1` or `stress2`) was more significantly effected by using adult coloring books.

`@hint`


`@pre_exercise_code`
```{r}
set.seed(1)
n=514
#Create rnorm function that allows for min and max
    rtnorm <- function(n, mean, sd, min = -Inf, max = Inf){
        qnorm(runif(n, pnorm(min, mean, sd), pnorm(max, mean, sd)), mean, sd)
    }
#Create rounding function that allows to round to numbers above 1
    mround <- function(x,base){ 
        base*round(x/base) 
    } 
#Dataframe
    Color<-data.frame(books=rbinom(n,1,.65))
    Color$age=round(rtnorm(n,mean=30,sd=5,min=25,max=60),0)
    Color$sex=ifelse(rbinom(n,1,.7)==1,"Female","Male")
    Color$stress1=round(rtnorm(n,mean=2,sd=3,min=1,max=10)+Color$age/15+(Color$sex=="Female")/4,0)
    Color$stress2=round(rtnorm(n,mean=1.5,sd=2,min=1,max=10)+Color$age/12+(Color$sex=="Female")/2 - Color$books/2,0)

```

`@sample_code`
```{r}
# 1) Estimate the immediate effect of coloring books, age, and gender on stress (i.e. the dependent variable should be stress1).

      Solution1<-glm()

# 2) Estimate the lagged effect of coloring books, age, and gender on stress (i.e. the dependent variable should be stress2)

      Solution2<-glm()

# 3) Which measurement of stress was more significantly effected by the `books` variable: stress1 or stress2? Answer with "stress1" or "stress2" To decide, compare the p-values of the `books` coefficient in Solution1 and Solution2.

      Solution3<-""

```

`@solution`
```{r}
Solution1<-glm(stress1~books+age+sex,data=Color)
Solution2<-glm(stress2~books+age+sex,data=Color)
summary(Solution1)
summary(Solution2)
Solution3<-"stress2"
```

`@sct`
```{r}
test_object("Solution3")
success_msg("Correct! Estimating an OLS model with a lagged dependent variable is pretty straightforward. While it is not the most sophisticated regression model with panel data, it is a great tool for measuring causal relationships where there is a delay between the cause and expected effect.")
```

---

## Individual Fixed Effects and Time Varying Treatments

```yaml
type: VideoExercise
key: 9f79358c00
lang: r
xp: 50
skills: 1
video_link: //player.vimeo.com/video/226207591
```


---

## Confounders in fixed effects models

```yaml
type: MultipleChoiceExercise
key: f5f1233e13
lang: r
xp: 50
skills: 1
```

Suppose we are interested in how exercising in a gym affects a person's self-esteem. We try to gather a random sample of individuals at multiple time points, and examine whether individuals who increased their frequency of exercise in a gym had higher self-esteem. However, when assessing the data, we realize that we oversampled white men, who are known to exercise at gyms at high rates, and who tend to have relatively high self-esteem. If we estimate a fixed effects model with this data, will the fact that we oversampled white men confound our results?

`@possible_answers`
- Yes
- No

`@hint`


`@pre_exercise_code`
```{r}

```

`@sct`
```{r}
msg1 = "Try again"
msg2 = "Correct! Our oversampling of men would clearly confound our estimates in a typical OLS regression model, but since we are only looking at within-person changes, the fact that our sample is biased towards a group of men who tend to exercise and have high self-esteem should not confound our results. As long as the effect of exercise on self-esteem is equal among white men and all other people, and that White men change their rate of exercise as often as the rest of the population, our fixed effects models with this data should accurately reflect the effect of exercise on self-esteem among all people. What is most important is how  the rate of exercise among our sample *changed* over time. If few individuals in our sample increased or decreased their rate of exercise, or if only particular groups tended to change their rates of exercise, our fixed effects models would be biased and have limited statistical power. Individuals whose behavior does not change between waves cannot add to our understanding about how a change in that behavior effects changes in their outcomes."
test_mc(correct = 2, feedback_msgs = c(msg1,msg2))
```

---

## The "Random Assignment" of Changes

```yaml
type: VideoExercise
key: f93e8e963c
lang: r
xp: 50
skills: 1
video_link: //player.vimeo.com/video/226207674
```


---

## Violating the random assignment of changes

```yaml
type: MultipleChoiceExercise
key: 09b18d2eae
lang: r
xp: 50
skills: 1
```

Daniel notices that whenever he buys a new WePhone, he feels incredibly happy. Daniel thinks that the world would be a much happier place if everyone owned a new WePhone, and is petitioning the government to subsidize the purchase of all new WePhones. To supplement his petition, Daniel surveys a group of  individuals at two time points. During both waves of the survey, Daniel asks the survey group how happy they are, and whether they own the WePhone 10S (the newest iteration of the WePhone). Daniel conducts a fixed effects model on this dataset, and finds a positive effect of owning a WePhone 10S on happiness (i.e. those who acquired a WePhone 10S in the subsequent wave appeared happier). Why might Daniel's survey be biased?

`@possible_answers`
- Because the type of person that would buy a WePhone 10S (i.e. be treated) is more likely to be someone who would derive happiness from owning a WePhone 10S.  
- Because people who can afford to buy a WePhone 10S are wealthier and therefore should be happier.
- Because Daniel didn't control for whether buying any new phone would make them happier.
- Because Daniel is a socialist.

`@hint`


`@pre_exercise_code`
```{r}

```

`@sct`
```{r}
msg1 = "Correct! Whenever your treatment group is made of people who self-selected into a treatment, you are likely to see some bias. The validity of Daniel's survey rests on his ability to prove that a random sample of the population underwent treatment. In other words, Daniel would need to demonstrate that the individuals who had purchased WePhones between each wave were identical to those who had not purchased WePhones. However, it is pretty likely that people only purchase a WePhone if they expect to derive some sort of utility from the purchase."
msg2 = "This is not a possible source of bias. A fixed effect model only looks at within-person changes; even if people who ended up buying WePhones tended to be wealthier, and even if wealth tended to make people happier, this would not be a source of bias because we are already controlling for differences across individuals."
msg3 = "Even if buying a new phone of any sort makes a person happier, this does not confound or refute Daniel's hypothesis that buying a new WePhone makes people happier."
msg4 = "Try again"
test_mc(correct = 1, feedback_msgs = c(msg1,msg2,msg3,msg4))
```

---

## Nonlinear time trends

```yaml
type: MultipleChoiceExercise
key: 0fe070c3f3
lang: r
xp: 50
skills: 1
```

The popular online retailer, Congo.com, is conducting an experiment to determine whether changing the color scheme of the website's product pages results in higher rates of sales. Congo.com decides to alter the color scheme on the store page for one of its brands of fidget spinners, and use a difference in differences model to determine how the sales of that brand of fidget spinners compared to another popular brand of fidget spinners. However, the sales of fidget spinners during the period of analysis climbed and fell dramatically. The sales trends are indicated on the graph to the right.

Based on this graph and your knowledge about difference in differences models, how did changing the website's color scheme appear to affect sales?

`@possible_answers`
- Changing the website's color scheme (treatment) increased sales.
- Changing the website's color scheme (treatment) decreased sales.

`@hint`


`@pre_exercise_code`
```{r}
    library(ggplot2)
    
    df<-data.frame(Month=factor(rep(c("March","April","May","June","July","August","September"),2)))
    df$Treated<-as.factor(rep(c("Yes","No"),each=7))
    df$Sales<-c(18.6,35.5,53,62.6,56,42.2,23.5,28.6,50.7,72,84,82,72,54)
    df$Month<-factor(df$Month, levels = c("March","April","May","June","July","August","September"))
    p<-ggplot(data=df,aes(x=Month,y=Sales,group=Treated,col=Treated))
    p+geom_line()+geom_point()+
      theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
            panel.background = element_blank(), axis.line = element_line(colour = "black"))+
      ggtitle("Fidget Spinner Sales by Experimental Group")+
      labs(y="Sales (in thousands)",color="Changed Color Scheme")
```

`@sct`
```{r}
msg1 = "Try again"
msg2 = "Correct! We could verify this answer with a difference in differences model. However, it is important to note that nonlinear trends like these can cause strange results in difference in differences models. We should be very cautious when modeling nonlinear trends."
test_mc(correct = 2, feedback_msgs = c(msg1,msg2))
```

---

## Does Posting Calorie Counts Lower Calorie Consumption? Causal Inference Bootcamp

```yaml
type: VideoExercise
key: 8a33d08193
lang: r
xp: 50
skills: 1
video_link: //player.vimeo.com/video/226207727
```


---

## Attrition in Panel Data

```yaml
type: MultipleChoiceExercise
key: 280768b7d2
lang: r
xp: 50
skills: 1
```

Attrition is an inevitability when conducting longitudinal surveys. Many survey respondents refuse to participate in a follow up survey, either to a respondent's lack of interest, or because it is no longer convenient for him to participate in the survey. However, attrition can cause bias. Imagine we conduct a survey about how restaurant reviews effected their number of diners. We gather information from restaurants at two time points, but only 70% of our initial sample participated in our follow up survey. In which of the following circumstances would this attrition be the most cause for concern (i.e. bias our estimates)?

`@possible_answers`
- If the Restaurants that failed to follow up tended to have worse reviews.
- If the Restaurants that failed to follow up tended to be part of chains.
- If the Restaurants that failed to follow up tended to be vegetarian.

`@hint`


`@pre_exercise_code`
```{r}

```

`@sct`
```{r}
msg1 = "Correct. While attrition across all of these traits could cause biases in our estimates, we would be most concerned by non-response that selected our sample on a trait that is related to our variables of interest. Not only does attrition make our sample less representative of the population of restaurants, it might bias our estimates substantially. In this case, what if it turned out that the reason that restaurants that had received worse reviews across the survey were less likely to respond in the second wave is because they went out of business? In this case, our survey would fail to capture information from the restaurants that were most significantly affected by their poor reviews. We would see a much smaller effect from having poor reviews on the number of diners that a restaurant has than if we included those restaurants that failed."
msg2 = "Try again"
msg3 = "Try again"
test_mc(correct = 1, feedback_msgs = c(msg1,msg2,msg3))
```

---

## Extreme Differences in Control and Treatment

```yaml
type: MultipleChoiceExercise
key: ce0e698f11
lang: r
xp: 50
skills: 1
```

The mayor of Gettysburg, Alaska is concerned about lead in the town's water. He is considering upgrading the town's water filters, but the city council requires him to provide evidence of the filters' effectiveness in other towns first. The Mayor decides to present a table (illustrated below) showing the amount of lead in the average Gettysburg household's water supply in the past decade (measured in parts per billion) versus the amount of lead in households of the neighboring town, Bristol, which had implemented this filtration system. The mayor also presents a difference in differences model of this data, which shows a significantly signifcant effect of the filtration system in Bristol. Is this data appropriate for conducting a difference in differences model? 


| Town        |   2000   |   2005   |   2010   |
|-------------|---------:|---------:|---------:|
| Gettysburg  |   21 ppb |   23 ppb |   22 ppb |
| Bristol     |13200 ppb | 3700 ppb |  200 ppb |

`@possible_answers`
- Yes
- No

`@hint`


`@pre_exercise_code`
```{r}

```

`@sct`
```{r}
msg1 = "Correct! In this case, the common trends assumption would hold; however, a difference in differences model would violate our assumption that the treatment has a linear effect on our outcome. If the treatment had a linear effect, we would expect implementing a water filter to have a negative amount in the amount of lead in its water supply. While we could try log transforming the data before conducting a difference in differences model, it is still uncertain whether the filtration system would have any effect in a town with such low baseline quantities of led in its water. They mayor of Gettysburg could make a much stronger argument if he found a comparison group that had similar baseline quantities of lead to Gettysburg"
msg2 = "Try again"
test_mc(correct = 1, feedback_msgs = c(msg1,msg2))
```

---

## Switchers

```yaml
type: VideoExercise
key: 7ba7ef8e01
lang: r
xp: 50
skills: 1
video_link: //player.vimeo.com/video/226207860
```


---

## Practice Running a Fixed and Random Effects Regression Models

```yaml
type: NormalExercise
key: e9cecd77c0
lang: r
xp: 100
skills: 1
```

Many students at Perry Elementary school write with gel pens. Iris, the school's librarian, is convinced that students could write faster if all the students were required to use ballpoint pens. To test her theory, Iris gets permission to conduct a secondary analysis of an annual student written exam. 

At the end of each school year, students are required to write a three paragraph answer to a short prompt. When students submit their exams, teachers note how long the students took to submit their exams, how many words were written in their exams, and ask the students how many hours of sleep they had slept the previous night. In addition, Iris identified whether the exams were written with a gel or ballpoint pen.

These style exams are given to students three years in a row. Iris was given permission to look at all three of these exams for a single class of students. Since it is possible that students who write with gel pens have different innate writing abilities than those who write with ballpoint pens, Iris decides to only study changes in writing speed for people who switched between using gel pens to ballpoint pens. 

Using the dataset, `Exams`, help Iris estimate the effect of using a ballpoint pen on a student's writing speed with *pooled OLS*, *fixed* and *random* effects models.

`@instructions`
- 1) Examine the structure of the data and the types of variables we have.
- 2) Look more closely at the values of our treatment variables.
- 3) Construct a pooled OLS model that measures the effect of ballpoint pens on student's writing speeds, controlling for age, sex, and hours slept (variables: `pen`,`speed`,`age`,`sex`,`sleep`) .
- 4) Examine the results on that model and interpret what they are saying.
- 5) Construct a fixed effects model that measures the effect of ballpoint pens on student's writing speeds, controlling for age, sex, and hours slept (variables: `pen`,`speed`,`age`,`sex`,`sleep`).
- 6) Determine whether the absolute value of the Gel pen effect is larger or smaller in the fixed effects model than in the pooled OLS model.
- 7) Construct a random effects model that measures the effect of ballpoint pens on student's writing speeds, controlling for age, sex, and hours slept (variables: `pen`,`speed`,`age`,`sex`,`sleep`).

`@hint`


`@pre_exercise_code`
```{r}
library(plm)
n=521
#Create rnorm function that allows for min and max
  rtnorm <- function(n, mean, sd, min = -Inf, max = Inf){
      qnorm(runif(n, pnorm(min, mean, sd), pnorm(max, mean, sd)), mean, sd)
  }
#Create rounding function that allows to round to numbers above 1
  mround <- function(x,base){ 
          base*round(x/base) 
  } 

set.seed(1)
#Dataframe
  Exams<-data.frame(id=1:n,wave=1)
#year 1:
    #age
        Exams$age<-10+rbinom(n=n,1,.2)
    #sex
        Exams$sex<-ifelse(rbinom(n=n,1,.55)==1,"Female","Male")
    #sleep
        Exams$sleep<-mround(rtnorm(n=n,mean=8,sd=1,min=5,max=9),.5)
    #pen
        Exams$pen<-ifelse(rbinom(n=n,1,.5+ifelse(Exams$sex=="Male",.25,0)+1/Exams$age)==1,"Ballpoint","Gel")
    #speed
        Exams$speed<-round((Exams$age/10+(Exams$sex=="Female")/3+(Exams$sleep/rtnorm(n,2,1,1,3)/4)+(Exams$pen=="Ballpoint")*rtnorm(n,.5,.1,0,1))*2+rnorm(n,3,.5),2)

#year 2:
  Exams2<-data.frame(id=1:n,wave=2)
    Exams2$age<-Exams$age+1
    Exams2$sex<-Exams$sex
    Exams2$sleep<-mround(Exams$sleep+rtnorm(n,0,.3,0,1),.5)
    Exams2$pen<-ifelse(rbinom(n,1,.75)==1,Exams$pen,ifelse(rbinom(n=n,1,.4+ifelse(Exams$sex=="Male",.25,0)+1/Exams$age)==1,"Ballpoint","Gel"))
    Exams2$speed<-round(((Exams2$age/10+(Exams2$sex=="Female")/3+(Exams2$sleep/rtnorm(n,2,1,1,3)/4)+(Exams2$pen=="Ballpoint")*rtnorm(n,.5,.1,0,1))*2+rnorm(n,3,.5))/5+Exams$speed,2)

#year 3:
Exams3<-data.frame(id=1:n,wave=3)
    Exams3$age<-Exams2$age+1
    Exams3$sex<-Exams2$sex
    Exams3$sleep<-mround(Exams2$sleep+rtnorm(n,0,.3,0,1),.5)
    Exams3$pen<-ifelse(rbinom(n,1,.75)==1,Exams2$pen,ifelse(rbinom(n=n,1,.4+ifelse(Exams2$sex=="Male",.25,0)+1/Exams2$age)==1,"Ballpoint","Gel"))
    Exams3$speed<-round(((Exams3$age/10+(Exams3$sex=="Female")/3+(Exams3$sleep/rtnorm(n,2,1,1,3)/4)+(Exams3$pen=="Ballpoint")*rtnorm(n,.5,.1,0,1))*2+rnorm(n,3,.5))/5+Exams2$speed,2)
  
#rbind
    Exams<-rbind(Exams,Exams2)
    Exams<-rbind(Exams,Exams3)
    Exams<-Exams[order(Exams$id,Exams$wave),]

```

`@sample_code`
```{r}
# 1) Before running any models, let's examine the data. The dataset contains seven variables. Wave refers to which date in the time series that we are looking at. Wave 1 refers to the first year the data was collected, whereas wave 3 refers to the last year that the data was collected. Since data was collected three times for each individual, each id variable occurs in the data set 3 times.

    str(Exams)
    head(Exams)

# 2) The meaning of the other variables is pretty intuitive. Sleep refers to the number of hours that the students slept the previous night, and speed refers to the number of words that each student wrote per a minute.

    summary(Exams$sleep)
    summary(Exams$speed)

# 3) Run a typical OLS model on the dataset, treating each repeated observation as an independent person (i.e. pooling the data), where speed is the dependent variable, pen is the independent variable, and where we have control variables for age, sex, and sleep. For reference, the syntax for glm models is glm(y ~ x + z1 + z2 + z3, data=df), where y is the dependent variable, x is the independent variable, z1-z3 are control variables, and df is the data frame. You do not need to use the id or wave variables in this model.

    Solution3<-glm()
  
# 4) If you answered question 3 correctly, you should see a large negative effect from using gel pens when you view the summary results: 

    summary(Solution3)
    
# However, when we pool the data this way, we are treating repeated observations as separate people. This is not true. Pooling our data this way will bias our parameter estimates, especially if there is variation in each student's underlying writing ability and these variations are correlated with whether they use gel pens.    
    
# 5) Now let's try estimating a fixed effects model. Random effects models work similarly to OLS models, except that it also controls for how observations in our sample are related. Basically, we want to control for student's different latent writing speeds, and only look at how switching between a gel and ballpoint pen effected their writing speeds compared to those who did not switch pens, To estimate a fixed effects model, we need to use the plm function. The syntax is  identical to glm, except that after the data statement, we specify how the data are grouped (by id and wave) and what type of effect that we want to examine (within person changes). Fill in the front part of the plm model so that it estimates the effect of using a gel pen on writing speed while controlling for age, sex, and sleep.
    
    Solution5<-plm(,index=c("id", "wave"), model="within")

# 6) If you answered question 5 correctly, you should notice that the model only gave us parameter estimates for sleep and gel pens. This is because the sex of students did not change across any years (again, we are only looking at how changes within individuals affected their writing speeds), and because the change in age was uniform across years for all respondents (that is, there was no variation in how student's ages changed across individuals - everyone grew one year older each wave). You should also notice that the parameter estimate for gel pens differs from the first model. Is the absolute value of the effect of gel pens larger or smaller in solution2 than in solution1. Answer with "Larger" or "Smaller".

    Solution6<-""

# 7 ) Although fixed effects models are very reliable, they often require large sample sizes to find statistically significant results, and they often exclude estimating effect that would be interesting to report in a regression model (for example, the effect of age and sex on writing speed). In such cases, we often use random effects models instead. Random effects models are almost identical to fixed effects models, except they make the somewhat strong assumption that the individual specific effects are uncorrelated with your other parameters. In this example, a random effects model assumes that every individual has a latent writing speed that is uncorrelated with anything else. Factors like age and sex simply improve their writing speeds. Estimate a random effects model that groups individuals by id and wave, but that also controls for age, sex, and sleep. The syntax should be identical to that in Question 2, except that the "model" parameter in the plm function should be switched from "within" to "random".

      Solution7<-plm()

```

`@solution`
```{r}
#models    
    Solution3<-glm(speed~pen+sleep+sex+age,data=Exams)
    Solution5<-plm(speed ~ age + sex + sleep + pen, data=Exams, index=c("id", "wave"), model="within")
    Solution6<-"Larger"
    Solution7<-plm(speed ~ age + sex + sleep + pen, data=Exams, index=c("id", "wave"), model="random")
    
```

`@sct`
```{r}
test_object("Solution7")
success_msg("Good work! You now know the basics to estimating fixed and random effects models with panel data. As you can see, these models often give very different results than typical pooled OLS models, and their estimates tend to be much more robust. You may still be wondering when you should use random versus fixed effects models. In general, if you can find statistically significant effects with fixed effects models, they are preferable to random effects models. However, there are a variety of tools available for comparing whether you get any benefit from using a fixed effects model (e.g. the Hausman test). In general, use you intution and see what other people have done in similar situations!")
```

---

## Putting It All Together: Electric Cars and Caring About Nature

```yaml
type: NormalExercise
key: 087c0debd1
lang: r
xp: 100
skills: 1
```

Located in a part of the country that has been losing heavy industry jobs for the last few decades, the Square Cities area includes the four cities of Greenville, Tarboro, Blackwater, and Milltown. While there are still a few factories and other industries in the area, these four city councils have been trying to make the area more attractive for new businesses and young millennials through new environmental policies, even though this is not always fully supported by the local residents who tend to be politically conservative. Nonetheless, the city of Greenville liked the idea of stimulating the purchase of electric cars, and in 2013, the city's governance decided to subsidize electric car purchases of Greenville residents. 

The city of Greenville now asks you to determine whether the city subsidy had a causal effect on the car purchase behavior of the town's residents. Let's use our analytical skills on panel data to complete the following:

`@instructions`
- 1) Estimate the association between qualifying for a car subsidy and purchasing an electric car with an OLS model on one cross section of the provided dataset (`survey2015`).
- 2) Estimate the association between changes in qualifying for a car subsidy and purchasing an electric car with a fixed effects model and panel data (`survey13.15lf`).
- 3) Estimate the association between qualifying for a car subsidy and purchasing an electric car with a lagged dependent variable approach and panel data (`survey13.15.wf`).
- 4) State whether the time trends reported in the provided data suggest that Greenville's subsidy was effective.
- 5) Determine whether Greenville and its neighbor, Blackwater, are relatively similar across demographic traits.
- 6) Compare trends of electric car ownership to those in Blackwater..


`@hint`


`@pre_exercise_code`
```{r}
# Authors: James Speckart, Brian Aronson, and Attila Gyetvai
library(ggplot2)
library(plyr)
library(plm)
# Initialization
nG <- 390
nB <- 403
n <- nG + nB
set.seed(100)

# City demographics
citydemos<-matrix(c("24.7k","48.6k",38.0,50.9,35.1,"43.8k", "23.1k","46.1k",35.9,51.7,34.8,"42.1k","24.3k","49.1k",38.2,50.4,34.9,"42.0k","25.3k","48.9k",37.8,50.8,35.2,"41.9k"), ncol=6, byrow=TRUE)
colnames(citydemos)<-c("Population","  Median Income", "  Median Age","  % Female", "  % Trees", "  City GDP per Capita")
rownames(citydemos)<-c("Greenville", "Blackwater", "Tarboro", "Milltown")
citydemos<-citydemos[1:2,]
citydemos<-citydemos[,-5]
citydemos<-as.table(citydemos)

# Set up data frame
# Year 1
survey1 <- data.frame(ID = 1:n, year = 2010, city = c(rep("Greenville", nG), rep("Blackwater", nB)))
survey1$income <- ifelse(survey1$city=="Greenville", 1000*(exp(rlnorm(nG, meanlog=1, sdlog=0.2))+30), 
                                                     1000*(exp(rlnorm(nB, meanlog=0.8, sdlog=0.2))+20))
incomeG <- survey1$income[survey1$city=="Greenville"]
incomeB <- survey1$income[survey1$city=="Blackwater"]
survey1$carenature <- 0.75*survey1$income + rnorm(n, 0, 20000)
survey1$carenature <- (survey1$carenature - min(survey1$carenature))/max(survey1$carenature - min(survey1$carenature))
carenatureG <- survey1$carenature[survey1$city=="Greenville"]
carenatureB <- survey1$carenature[survey1$city=="Blackwater"]
cartypeG <- ifelse(incomeG > quantile(incomeG, 0.25), rbinom(nG, 1, 0.9), rbinom(nG, 1, 0.3))
cartypeG[cartypeG>0] <- 1 + rbinom(length(carenatureG[cartypeG>0]), 1, 0.5*carenatureG[cartypeG>0]/max(carenatureG[cartypeG>0]))
cartypeB <- ifelse(incomeB > quantile(incomeB, 0.25), rbinom(nB, 1, 0.9), rbinom(nB, 1, 0.3))
cartypeB[cartypeB>0] <- 1 + rbinom(length(carenatureB[cartypeB>0]), 1, 0.35*carenatureB[cartypeB>0]/max(carenatureB[cartypeB>0]))
survey1$cartype <- c(cartypeG, cartypeB)
survey1$cartype <- factor(survey1$cartype, labels = c("None", "Combustion engine", "Electric"))

# From year 2 on
datagen <- function(df.old, year.var, trans.mat.G, trans.mat.B) {
  
  df.new <- data.frame(ID = 1:n, year = year.var, city = c(rep("Greenville", nG), rep("Blackwater", nB)))
  df.new$income <- df.old$income * (1.02 + rnorm(n, 0, 0.05))
  df.new$carenature <- pmin(pmax(df.old$carenature + rnorm(n, 0, 0.05), 0), 1)
  df.new$cartype[df.new$city=="Greenville" & df.old$cartype=="None"] <- which(rmultinom(length(which(df.new$city=="Greenville" & df.old$cartype=="None")), 1, trans.mat.G[1,])==1, arr.ind = TRUE)[,1] - 1
  df.new$cartype[df.new$city=="Blackwater" & df.old$cartype=="None"] <- which(rmultinom(length(which(df.new$city=="Blackwater" & df.old$cartype=="None")), 1, trans.mat.B[1,])==1, arr.ind = TRUE)[,1] - 1
  df.new$cartype[df.new$city=="Greenville" & df.old$cartype=="Combustion engine"] <- which(rmultinom(length(which(df.new$city=="Greenville" & df.old$cartype=="Combustion engine")), 1, trans.mat.G[2,])==1, arr.ind = TRUE)[,1] - 1
  df.new$cartype[df.new$city=="Blackwater" & df.old$cartype=="Combustion engine"] <- which(rmultinom(length(which(df.new$city=="Blackwater" & df.old$cartype=="Combustion engine")), 1, trans.mat.B[2,])==1, arr.ind = TRUE)[,1] - 1
  df.new$cartype[df.new$city=="Greenville" & df.old$cartype=="Electric"] <- which(rmultinom(length(which(df.new$city=="Greenville" & df.old$cartype=="Electric")), 1, trans.mat.G[3,])==1, arr.ind = TRUE)[,1] - 1
  df.new$cartype[df.new$city=="Blackwater" & df.old$cartype=="Electric"] <- which(rmultinom(length(which(df.new$city=="Blackwater" & df.old$cartype=="Electric")), 1, trans.mat.B[3,])==1, arr.ind = TRUE)[,1] - 1
  df.new$cartype <- factor(df.new$cartype, labels = c("None", "Combustion engine", "Electric"))
  return(df.new)
                      
}
trans.mat.pre <- matrix(c(0.8, 0.15, 0.05, 0.075, 0.8, 0.125, 0.05, 0.075, 0.875), 3, 3, byrow = TRUE)
trans.mat.post <- matrix(c(0.8, 0.1, 0.1, 0.1, 0.65, 0.25, 0.05, 0.06, 0.89), 3, 3, byrow = TRUE)
survey2 <- datagen(survey1, 2011, trans.mat.pre, trans.mat.pre)
survey3 <- datagen(survey2, 2012, trans.mat.pre, trans.mat.pre)
survey4 <- datagen(survey3, 2013, trans.mat.post, trans.mat.pre)
survey5 <- datagen(survey4, 2014, trans.mat.post, trans.mat.pre)
survey6 <- datagen(survey5, 2015, trans.mat.post, trans.mat.pre)
survey7 <- datagen(survey6, 2016, trans.mat.post, trans.mat.pre)

# Long data
survey <- rbind(survey1, survey2, survey3, survey4, survey5, survey6, survey7)
survey <- survey[order(survey$ID, survey$year),]

# Lag carenature
# lg <- function(x)c(NA, x[1:(length(x)-1)])
# survey <- ddply(survey, ~ID, transform, l.carenature = lg(carenature))

# DD plot
# df.plot <- aggregate(cartype=="Electric" ~ year + city, survey, sum)
# df.tmp <- aggregate(cartype!="None" ~ year + city, survey, sum)
# names(df.plot) <- c("year", "city", "num.electric")
# df.plot$share.electric <- df.plot$num.electric/df.tmp$cartype
# ggplot(df.plot[df.plot$city=="Greenville",], aes(x = year, y = share.electric, color = city)) +
#   geom_point() +
#   geom_line() +
#   labs(x = "Year", y = "Share of electric cars")
# ggplot(df.plot, aes(x = year, y = share.electric, color = city)) +
#   geom_point() +
#   geom_line() +
#   labs(x = "Year", y = "Share of electric cars")

#Converting DD Plot into single command to run in sample code
GraphRates<-function(){
df.plot <- aggregate(cartype=="Electric" ~ year + city, survey, sum)
df.tmp <- aggregate(cartype!="None" ~ year + city, survey, sum)
names(df.plot) <- c("year", "city", "num.electric")
df.plot$share.electric <- (df.plot$num.electric/df.tmp$cartype)/5
ggplot(df.plot[df.plot$city=="Greenville",], aes(x = year, y = share.electric, color = city)) +
    geom_point() +
    geom_line() +
    labs(x = "Year", y = "Share of electric cars")
}

GraphComparingCities<-function(){
df.plot <- aggregate(cartype=="Electric" ~ year + city, survey, sum)
df.tmp <- aggregate(cartype!="None" ~ year + city, survey, sum)
names(df.plot) <- c("year", "city", "num.electric")
df.plot$share.electric <- (df.plot$num.electric/df.tmp$cartype)/5
ggplot(df.plot[df.plot$city=="Greenville",], aes(x = year, y = share.electric, color = city)) +
    geom_point() +
    geom_line() +
    labs(x = "Year", y = "Share of electric cars")
ggplot(df.plot, aes(x = year, y = share.electric, color = city)) +
    geom_point() +
    geom_line() +
    labs(x = "Year", y = "Share of electric cars")
}



#Beginning parts of question:
rtnorm <- function(n, mean, sd, min = -Inf, max = Inf){
        qnorm(runif(n, pnorm(min, mean, sd), pnorm(max, mean, sd)), mean, sd)
    }
mround <- function(x,base){
    base*round(x/base)
}
set.seed(100)
n=213
#prep initial data
survey2015<-data.frame(
    below.threshold=rbinom(n,1,.45),
    college=rbinom(n,1,.25),
    male=rbinom(n,1,.6),
    id=1:n
)
#allow threshold info to vary
    survey2013<-survey2015
    subsam<-rbinom(n,1,.2)
    survey2013$below.threshold[subsam==1]<-1-survey2013$below.threshold[subsam==1]
#make miles driven correlated with male and threshold
    survey2013$milesdriven<-rtnorm(n,mean=15000,sd=8000,min=6000,max=25000)+4000*survey2013$male-2000*survey2013$below.threshold
#determine factors for buying electric
    a<-survey2013$milesdriven/10000+.4*survey2013$male+.05*survey2013$college+survey2013$below.threshold
a<-(a-min(a))/max(a)/1.5
summary(a)
    survey2013$bought.electric<-rbinom(n,1,.05)
    survey2015$bought.electric<-rbinom(n,1,a)
#miles driven in 2015 determined by buying electric
    survey2015$milesdriven<-survey2013$milesdriven+
    rtnorm(n,mean=0,sd=2000,min=-5000,max=5000)
    survey2015$milesdriven[survey2015$bought.electric==1]<-survey2015$milesdriven[survey2015$bought.electric==1]*  rtnorm(length(survey2015$milesdriven[survey2015$bought.electric==1]),mean=1.5,sd=.1,min=1,max=2)
survey2015$milesdriven[survey2015$milesdriven>30000]<-survey2015$milesdriven[survey2015$milesdriven>30000]-10000

#two wave versions of this (long format)
    survey13.15lf<-rbind(survey2013,survey2015)
    survey13.15lf$year<-rep(c(2013,2015),each=n)
    survey13.15lf<-survey13.15lf[order(survey13.15lf$id,survey13.15lf$year),]

#two wave versions of this (wide format)
    temp1<-survey2013
    names(temp1)<-paste(names(temp1),"2013",sep=".")
    temp2<-survey2015
    names(temp2)<-paste(names(temp2),"2015",sep=".")
    survey13.15wf<-cbind(temp1,temp2)


```

`@sample_code`
```{r}
# Intro: Greenville collected a lot of relevant information about their experiment. As a first cut, they shared a series of annual surveys they collect about town-member's demographic's and car-purchases. Greenville suggests that you compare the rate of electric car purchases in 2015 for citizens who were below the necessary income threshold for attaining a subsidy to citizens who were above the necessary income threshold for attaining a subsidy (i.e. those who did not qualify). They expect that individuals right below the income threshold would be more likely to purchase electric cars, controlling for other demographic characteristics. 

# 1) Use dataset "survey2015" to estimate the association between qualifying for a car purchase (`below.threshold`) and purchasing an electric car (`bought.electric`) controlling for college education (`college`), gender (`male`), and miles driven that year (`milesdriven`)

    summary(survey2015)
    glm()

# Note: For simplicity, we ask that you run this as an OLS regression (glm's default), rather than as logistic regression. However, in most cases, a logistic regression would be more appropriate for this question.
    
# 2) If you had answered the above correctly, you will see that being below the threshold for subsidies had no significant association with buying an electric car. Why might be the case? Greenville argues that our model is misspecified, as our independent variables might be endogenously confounded. Greenville suggests that we try running a fixed effects model, to more directly test whether falling below or above the subsidy threshold over time motivates households to purchase electric cars. Run a fixed effects model on a the long-formatted panel dataset `survey13.15lf` to see whether people who qualified for subsidies (`below.threshold`) in 2015 but not 2013 (or vice-versa) were more likely to purchase electric cars (`bought.electric`), controlling for (`milesdriven`). As a hint, we filled in some of the model for you.
  
    summary(survey13.15lf)
    plm( index=c("id", "year"), model="within")    
    
# Note: Notice that we did not control for college or male in this model. The reason why is because these variables did not change (i.e. they were time-invariant); if we include them in a fixed effects model, we will not estimate a coefficient for them, as they have no variance in the fixed-effects approach.
    
# 3) The answer still seems to be "no", the subsidy did not incentivize car purchases. Greenville now argues that the problem with the fixed-effects approach is that their sample was not large enough. Less than a quarter of the sample ever changed income thresholds. Moreover, the causal relationship between miles driven and buying an electric car is not clear in this model; driving more miles may incentivize electric car ownership, but buying an electric car may incentivize driving more miles. This unclear causal relationship may have confounded the relationship between qualifying for subsidies and purchasing an electric car. As a final test, Greenville suggests that you run a regression model with a lagged dependent variable. Specifically, they ask you use a wide-format version of this panel data (survey13.15wf) to test the effect of qualifying for a car purchase in 2013 (`below.threshold.2013`) on purchasing an electric car in 2015 (`bought.electric.2015`) controlling for other 2013 variables, including college education (`college.2013`), gender (`male.2013`), and miles driven that year (`milesdriven.2013`)
    
    summary(survey13.15wf)
    glm()
     
# Note: If you ran the above model correctly, you will see a positive statistically significant effect of qualifying for electric car subsidies in 2013 on electric car ownership in 2015. It seems that the potentially reverse causal relationship between miles driven and electric car ownership in 2015 was confounding this relationship. 

# 4a) The mayor of Greenville reflects on our findings, and now asks and important question: "The subsidies seem to motivate people near the subsidy threshold to purchase more electric cars, but does this really matter in the grand scheme of things? Did the policy really improve rates of electric car ownership in Greenville?" One of his consultants sugggests that we look at the following graph (to see it, run GraphRates()) to generate a graph that displays the share of electric cars within all cars in Greenville and neighboring Blackwater on a year-by-year basis, including when Greenville created a subsidy for electric cars in 2013: 

    GraphRates()      

# 4b) This consultant argues that rates of electric car ownership have grown a great deal over the past few years, therefore the subsidy must have been effective. Based on this graph, can you tell whether the subsidy was effective? Write "yes" or "no" for Solution1.
    
    Solution4 <- ""

# 5) Let's think of a way to use the Difference-in-Differences approach to find out if this is a causal effect. In the Difference-in-Differences approach, we should try to compare rates of electric car ownership among otherwise very similar cities. The mayor of Greenville suggests we compare Greenville to the nearby city of Blackwater, which did not subsidize electric car purchases over the same years. If the cities are very similar across key demographics, then we can feel confident that we can compare them with a difference-in-differences approach. Print out the table of collected demographics with the following line of code, and answer the following question with "yes" or "no": Would you consider the cities to be relatively similar across these core demographic traits?

    print(citydemos, right=TRUE)
    Solution5 <- ""

# Note: We accept all solutions for the above question, as the does not have a clear answer. Keep in mind that these are very much the type of question you will encounter in the real world.
    
# 6) Regardless of your answer above, let's assume that the two cities are similar enough to warrant comparison. Run the following code to illustrate a comparison of the share of electric cars owned in Greenville and Blackwater. Assuming the differences are statistically significant, can you say with relative confidence whether the subsidy worked in Greenville? Write "yes" or "no" for Solution2.
    
    GraphComparingCities()
    Solution6 <- ""

    
```

`@solution`
```{r}
#1)
summary(glm(bought.electric~below.threshold+college+male+milesdriven,data=survey2015))
#2)
summary(plm(bought.electric~below.threshold+college+male+milesdriven, data = survey13.15lf, index=c("id", "year"), model="within"))
#3)
summary(glm(bought.electric.2015~below.threshold.2013+college.2013+male.2013+milesdriven.2013,data=survey13.15wf))
Solution4 <- "no"
Solution5 <- "maybe"
Solution6 <- "yes"
```

`@sct`
```{r}
test_object("Solution4")
test_object("Solution6")
success_msg("Congratulations! By now, you should have a sense of the sorts of causal inference problems that panel data can help us overcome. There are a variety of methodological approaches you can take with panel data, each with their own pros and cons. This is why it is always critical to carefully think through what your cause and outcomes of interest are, and how they might be related causally. If you found this module useful, consider trying all of ModU Duke's modules on causal inference.")
```
